---
title: "MALF of the Insula on an ABIDE patient"
author: "John Muschelli"
date: '`r Sys.Date()`'
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
bibliography: ../refs.bib      
---

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/malf_insula/index.R).


```{r setup, include=FALSE}
library(methods)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, comment = "")
```

# Package Version

This tutorial requires the `hcp` package (>= 0.5):

```{r}
ver = installed.packages()["hcp", "Version"]
if (compareVersion(ver, "0.5") < 0) {
  stop(paste0("Need to update hcp, ", 
              "devtools::install_github('muschellij2/hcp')")
  )
}
```
# Data

We will be using the data from the Autism Brain Imaging Data Exchange [ABIDE](http://fcon_1000.projects.nitrc.org/indi/abide/) [@di2014autism].  The data consists of a children with and without autism from the 1000 Functional Connectomes Project and INDI. I chose a child that was young (around 8 years old) so that we can see if this method works for younger brains:

```{r get_data}
library(hcp)
fcp_data = download_hcp_file(
    paste0("data/Projects/ABIDE/RawData/", 
        "KKI/0050784/session_1/anat_1/",
        "mprage.nii.gz"),
    bucket = "fcp-indi",
    sign = FALSE)
print(fcp_data)
```

Note, you do not need an API key for this data.  We have an MPRAGE for this child downloaded to the disk.

## Visualization

```{r orig_ortho}
library(neurobase)
std_img = readnii(fcp_data)
ortho2(std_img)
```

## Reorientation
Although this image seems to be in RPI orientation, we can check using `getForms`:

```{r forms}
library(fslr)
getForms(fcp_data)[c("ssor", "sqor")]
```

This image is indeed in RPI, but we may want to ensure this is the case an use the `rpi_orient` function from fslr so that each image would be in the same orientation:

```{r reor}
reor = rpi_orient(fcp_data)
img = reor$img
```

## Brain Extraction

Here we will use `fslbet_robust` to skull strip the image.  The "robust"
```{r bet}
library(extrantsr)
bet = fslbet_robust(img, swapdim = FALSE)
```

```{r plot_bet}
ortho2(robust_window(img), bet)
```

```{r obet}
rb = robust_window(img)
bet2 = fslbet_robust(rb, swapdim = FALSE)
```
```{r obet2}
ortho2(robust_window(img), bet2)
```

### Zooming in 

Now that we have only the brain image, we can drop extraneous dimensions.  This dropping is for visualization, but also we may not want these extra dimensions affecting any registration.  

```{r dropping}
dd = dropEmptyImageDimensions(bet > 0,
    other.imgs = bet)
run_img = dd$other.imgs
```

```{r orun}
ortho2(run_img)
ortho2(robust_window(run_img))
```


## Labeled data

```{r get_labs, eval = FALSE}
root_template_dir = file.path(
    "/dcl01/smart/data", 
    "structural", 
    "Templates", 
    "MICCAI-2012-Multi-Atlas-Challenge-Data")
template_dir = file.path(root_template_dir,
    "all-images")
```

```{r labs, eval = FALSE}
library(readr)
labs = read_csv(file.path(root_template_dir,
    "MICCAI-Challenge-2012-Label-Information.csv"))

niis = list.files(
    path = template_dir,
    pattern = ".nii.gz", 
    full.names = TRUE)
bases = nii.stub(niis, bn = TRUE)
templates = niis[grep("_3$", bases)]

df = data.frame(
    template = templates,
    stringsAsFactors = FALSE)
df$ss_template = paste0(
    nii.stub(df$template),
    "_SS.nii.gz")
df$label = paste0(
    nii.stub(df$template),
    "_glm.nii.gz")
stopifnot(all(file.exists(unlist(df))))

indices = labs[grep("sula", tolower(labs$name)),]
indices = indices$label
```

```{r reading_data, eval = FALSE}
library(pbapply)
lab_list = pblapply(df$label,
    function(x) {
    img = fast_readnii(x)
    niftiarr(img, 
        img %in% indices
        )    
})

temp_list = pblapply(df$ss_template, readnii)

inds = 1:10
tlist = temp_list[inds]
llist = lab_list[inds]
```

```{r running_malf, eval = FALSE}
res = malf(infile = run_img,
    template.images = tlist,
    template.structs = llist,
    keep_images = FALSE,
    outfile = "test_malf_mprage_insula.nii.gz")

wimg = robust_window(run_img)
png("test_malf_image.png")
ortho2(wimg, xyz = xyz(res))
dev.off()


png("test_malf_image_overlay.png")
ortho2(wimg, res, xyz = xyz(res), 
    col.y = scales::alpha("red", 0.5))
dev.off()
```

