---
title: "Rxnat Paper Case Scenario"
author: "Adi Gherman"
date: "2020-06-22"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true      
bibliography: ../refs.bib      
---

All code for this document is located at [here]().

```{r setup, include=FALSE}
library(Rxnat)
library(dplyr)
library(fslr)
library(extrantsr)
library(malf.templates)
library(scales)
library(WhiteStripe)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, comment = "")
```


In this tutorial I will discuss performing tissue class segmentation using the `FAST` function from `FSL` and it's wrapper function in `fslr`, `fast_nobias` [@muschelli2015fslr]. 

# Data Packages
To perform this analysis, I will select one subject from the IXI neuroimaging dataset hosted on [https://nitrc.org/](https://nitrc.org/) [@kennedy2016nitrc]. The `Rxnat` package is required to be installed in order to access the NITRC image repository as well as an account with NITRC: [new account request](https://www.nitrc.org/account/register.php)
```{r, eval = FALSE}
packages = installed.packages()
packages = packages[, "Package"]
if (!"Rxnat" %in% packages) {
  source("https://neuroconductor.org/neurocLite.R")
  neuroc_install("Rxnat")    
}
```

# Loading Data
Please visit the [general Rxnat tutorial](../rxnat/index.html) on how to setup the Rxnat environment variables. Next we'll read and reorient the T1 image using the `readrpi` function from the `fslr` package.
```{r data, message = FALSE, warning = FALSE, cache = TRUE, fig.cap = "T1 Weighted Image"}
library(Rxnat)
nitrc <- xnat_connect("https://nitrc.org/ir", xnat_name="NITRC")
# Download the subject T1 weighted image
file_path <- nitrc$download_dir(
                experiment_ID = 'NITRC_IR_E10464',
                scan_type = "T1",
                extract = TRUE)
t1_fname <- file_path[1]
t1 <- readrpi(t1_fname)
ortho2(t1, add.orient = TRUE)
```

# Remove neck and drop empty dimensions
The neck removal step is implemented using the `remove_neck` function from the `extrantsr` package. The empty image dimensions (including the neck slices) can be dropped by using the function `dropEmptyImageDimensions` from the `neurobase` package.
```{r remove-neck, message = FALSE, warning = FALSE, cache = TRUE}
noneck = remove_neck(file_path,
    template.file = fslr::mni_fname(brain = TRUE, mm = 1),
    template.mask = fslr::mni_fname(mm = 1, brain = TRUE, mask = TRUE),
    verbose = FALSE
)
red = dropEmptyImageDimensions(noneck)
red <- readrpi(red)
ortho2(red, add.orient = TRUE)
```

# Inhomogeneity correction
To correct the bias field signal we use the `bias_correct` function from the `extrantsr` package, which uses the N4 inhomogeneity correction.
```{r correction, message = FALSE, warning = FALSE, cache = TRUE}
t1_n4 = bias_correct(red,
                     correction = "N4",
                     outfile = tempfile(fileext = ".nii.gz"), retimg = FALSE
)
t1_n4 <- readrpi(t1_n4)
```

# Malf registration
Once images are bias field corrected, we apply brain extraction using a form of multi-atlas label fusion (MALF). MALF uses a collection of previously labeled brain images (atlases), aligns the T1-weighted image to each atlas, and obtains a labeled T1-weighted image for each registration. This approach is implemented using the `malf` function from the `malf.templates` package, which includes the templates from the 2012 MICCAI Multi-Atlas Labeling Challenge.

```{r malf, message = FALSE, warning = FALSE, cache = TRUE}
timgs = mass_images(n_templates = 35)
ss = malf(infile = t1_n4,
          template.images = timgs$images,
          template.structs = timgs$masks,
          keep_images = FALSE,
          verbose = FALSE)
```

# Perform skull stripping
To do this we use the `preprocess_mri_within` function from the `extrantsr` package. This function performs N4 bias correction, image registration (if multi- sequence data is given), skull stripping (estimating the brain mask if one is not supplied), and brain mask application to the registered images.

```{r ss, message = FALSE, warning = FALSE, cache = TRUE}
proc_outfile <- paste0("T1_Processed.nii.gz")
proc_outfile <- file.path(tempdir(),proc_outfile)
skull_ss <- preprocess_mri_within(
  files = t1_n4,
  outfiles = proc_outfile,
  correction = "N4",
  maskfile = ss,
  correct_after_mask = FALSE)
t1_ss <- readrpi(proc_outfile)
ortho2(red,
       t1_ss,
       col.y=alpha("red", 0.3),
       add.orient = TRUE)
```

# Perform WhiteStripe intensity normalization
WhiteStripe intensity normalization can be implemented using the `whitestripe` and` whitestripe_norm` functions from the `WhiteStripe` package.
```{r ws, message = FALSE, warning = FALSE, cache = TRUE}
ind = whitestripe(img = t1_ss, type = "T1", stripped = TRUE)$whitestripe.ind
ws_t1 = whitestripe_norm(t1_ss, ind)
```

# Perform segmentation
This is implemented here using the FAST function from FSL. FAST segments a 3D brain image into different tissue types. We will use the `fast_nobias` from the `fslr` package, which assumes that the the bias field was removed.
```{r tcs, message = FALSE, warning = FALSE, cache = TRUE}
ss_tcs = fslr::fast_nobias(ws_t1,
                  verbose = TRUE)
double_ortho(ws_t1,
             ss_tcs,
             add.orient=TRUE)
```

# Session Info

```{r, cache = FALSE}
devtools::session_info()
```

# References
