---
title: "Skull Stripping CT data"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true      
---


All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/ss_ct/index.R).


```{r setup, include=FALSE}
library(dcm2niir)
library(ichseg)
library(dplyr)
library(fslr)
library(extrantsr)
library(TCIApathfinder)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, comment = "")
```


# Goal
In this tutorial, we will discuss skull-stripping (or brain-extracting) X-ray computed tomography (CT) scans.  We will use data from TCIA (http://www.cancerimagingarchive.net/) as there is a great package called [`TCIApathfinder`](https://CRAN.R-project.org/package=TCIApathfinder) to interface with TCIA.

## Using TCIApathfinder

In order to use `TCIApathfinder`, please see the [vignette to obtain API keys](https://cran.r-project.org/web/packages/TCIApathfinder/vignettes/introduction.html).  Here we will look at the collections:

```{r get_mods}
library(TCIApathfinder)
library(dplyr)
collections = get_collection_names()
collections = collections$collection_names
head(collections)

mods = get_modality_names(body_part = "BREAST")
head(mods$modalities)

```

## Getting Body Part Information
Here we can see all the parts of the body examined.

```{r get_body_part}
bp = get_body_part_names()
bp$body_parts
```

Particularly, these areas are of interest.  There seems to be a "bug" in `TCIApathfinder::get_series_info` which is acknowledged in the help file.  Namely, the `body_part_examined` is not always a parameter to be set.  We could get all the series info for all the collections from the code below, but it takes some times (> 15 minutes):
```{r bp_get, eval = FALSE}
# could look for any of these
get_bp = c("BRAIN", "HEAD", "HEADNECK")

# takes a long time
res = pbapply::pblapply(collections, function(collection) {
  x = get_series_info(
    collection = collection, 
    modality = "CT")
  x$series
})
```

## Getting Series

Here we will gather the series information for a study we know to have head CT data:

```{r series_info}
collection = "CPTAC-GBM"
series = get_series_info(
  collection = collection, 
  modality = "CT")
series = series$series
head(series)
```

Here we grab the first series ID from this data which has a description of "HEAD STD" for standard head:

```{r download_unzip, cache=FALSE}
std_head = series %>% 
  filter(grepl("HEAD STD", series_description))
series_instance_uid = std_head$series_instance_uid[1]

download_unzip_series = function(series_instance_uid,
                                 verbose = TRUE) {
  tdir = tempfile()
  dir.create(tdir, recursive = TRUE)
  tfile = tempfile(fileext = ".zip")
  tfile = basename(tfile)
  if (verbose) {
    message("Downloading Series")
  }
  res = save_image_series(
    series_instance_uid = series_instance_uid, 
    out_dir = tdir, 
    out_file_name = tfile)
  if (verbose) {
    message("Unzipping Series")
  }  
  stopifnot(file.exists(res$out_file))
  tdir = tempfile()
  dir.create(tdir, recursive = TRUE)
  res = unzip(zipfile = res$out_file, exdir = tdir)
  L = list(files = res,
           dirs = unique(dirname(normalizePath(res))))
  return(L)
}
# Download and unzip the image series

file_list = download_unzip_series(
  series_instance_uid = series_instance_uid)
```

## Converting DICOM to NIfTI

We will use [`dcm2niix`](https://github.com/rordenlab/dcm2niix) to convert from DICOM to NIfTI.  The function `dcm2niix` is wrapped in `dcm2niir`.  We will use `dcm2niir::dcm2nii` to convert the file.  We use `check_dcm2nii` to grab the relevant output files:
 
```{r dcm2nii, cache = FALSE}
library(dcm2niir)
dcm_result = dcm2nii(file_list$dirs)
result = check_dcm2nii(dcm_result)
```

Here we read the data into `R` into a `nifti` object:
```{r readnii, cache = FALSE}
library(neurobase)
img = readnii(result)
ortho2(img)
range(img)
```

Here we will use `neurobase::rescale_img` to make sure the minimum is $-1024$ and the maximum is $3071$.  The minimum can be lower for areas outside the field of view (FOV).  Here we plot the image and the Winsorized version to see the brain tissue:

```{r}
img = rescale_img(img, min.val = -1024, max.val = 3071)
ortho2(img)
ortho2(img, window = c(0, 100))
```

## Skull Strip

We can skull strip the image using `CT_Skull_Strip` or `CT_Skull_Stripper`.  The `CT_Skull_Stripper` has a simple switch to use `CT_Skull_Strip` or `CT_Skull_Strip_robust`.  
```{r ss}
library(ichseg)
ss = CT_Skull_Strip(img, verbose = FALSE)
ortho2(img, ss > 0, 
       window = c(0, 100),
       col.y = scales::alpha("red", 0.5))
```

The `CT_Skull_Strip_robust` function does 2 neck removals using `remove_neck` from `extrantsr` and then find the center of gravity (COG) twice to make sure the segmentation focuses on the head.  In some instances, the whole neck is included in the scan, such as some of the head-neck studies in TCIA.


# Showing a Robust Example with the neck

## Getting Series

Here we will gather the series information for the `Head-Neck Cetuximab` collection:

```{r robust_neck}
collection = "Head-Neck Cetuximab"
series = get_series_info(
  collection = collection, 
  modality = "CT")
series = series$series
whole_body = series %>% 
  filter(grepl("WB", series_description))
```


```{r download_unzip_robust, cache=FALSE}
file_list = download_unzip_series(
  series_instance_uid = series$series_instance_uid[1])
```

```{r dcm2nii_robust, cache=FALSE}
dcm_result = dcm2nii(file_list$dirs, merge_files = TRUE)
result = check_dcm2nii(dcm_result)
```

Here we see the original data has a lot of the neck and some of the shoulders in the scan:
```{r readnii2}
img = readnii(result)
img = rescale_img(img, min.val = -1024, max.val = 3071)
ortho2(img, window = c(0, 100))
```

We will try `CT_Skull_Strip` without adding any robust options:
```{r ss2}
ss_wb = CT_Skull_Strip(img, verbose = FALSE)
ortho2(ss_wb, window = c(0, 100))
```

We see that this does not work very well.  We will use the robust version.  Here we use `CT_Skull_Stripper`, which will call `CT_Skull_Strip_robust`.  This will run `extrantsr::remove_neck`, runs `CT_Skull_Strip`, then estimates a new center of gravity (COG) and then run `CT_Skull_Strip` again, and then run some hole filling:

```{r ss_robust}
ss_wb_robust = CT_Skull_Stripper(img, verbose = FALSE, robust = TRUE)
ortho2(ss_wb_robust, window = c(0, 100))
```

We see that this robust version works well for even data with the neck.  We can try it on a whole body image as well.  

# The website data
We could also look at the website, but these do not always correspond to the API and get all the necessary results.

```{r tabler}
library(rvest)
library(dplyr)
x = read_html("https://www.cancerimagingarchive.net/collections/")
tab = html_table(x)[[1]]
head_tab = tab %>% 
  filter(grepl("Head|Brain", Location),
         grepl("CT", `Image Types`), 
         Access == "Public")
brain_tab = tab %>% 
  filter(grepl("Brain", Location),
         grepl("CT", `Image Types`), 
         Access == "Public")
brain_tab
```

In `brain_tab`, we see we have a few collections.  We are going to use the collection `r collection` from above.

## Getting Patient Information

We could sample patients from the collection here and get the patient information:

```{r seeded}
set.seed(20181203)

patients = get_patient_info(collection = collection)
info = patients$patients
head(info)
```

Though we are not guaranteed the data will have Brain CT data.  We will use the `series` variable to grab a relevant scan.

