---
title: "DTI Analysis using rcamino for HCP data"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true      
---

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/DTI_analysis_rcamino_hcp/index.R).

```{r setup, include=FALSE, message=FALSE}
library(knitr)
library(methods)
library(neurobase)
knitr::opts_chunk$set(comment = "")
```

# Resources and Goals
Much of this work has been adapted by the Tutorial for DTI analysis from ISMRM 2015: [http://camino.cs.ucl.ac.uk/index.php?n=Tutorials.ISMRM2015](http://camino.cs.ucl.ac.uk/index.php?n=Tutorials.ISMRM2015).  We will show you a few steps that have been implemented in `rcamino`: `camino_pointset2scheme`, `camino_modelfit`, `camino_fa`, `camino_md`, and `camino_dteig`.  

# Data Location


# Reading in the Data
First, we download the data from HCP.  You must have your access keys set (see [Getting Data from the Human Connectome Project (HCP)](../neurohcp/index.html)).  

We will use the `neurohcp` package to download one subject data.

```{r, echo = TRUE}
library(neurohcp)
hcp_id = "100307"
r = download_hcp_dir(paste0("HCP/", hcp_id, "/T1w/Diffusion"))
print(basename(r$output_files))
```

It contains 4 files:

1.  `data.nii.gz` - a 4D image of the DWI data.
2.  `nodif_brain_mask.nii.gz` - A brain mask of the DTI data
3.  `bvals` - a text file with the b-values
4.  `bvecs` - a text file with the b-vectors as the first 3 columns. 


## Creating 


As `dtifit` requires the b-values and b-vectors to be separated, and this data has b-values of $1000$ when the b-vectors is not zero.  **This is very important and you must know where your b-values and b-vectors are when doing your analyses and what units they are in.**  


```{r bvecs}
library(rcamino)
camino_set_heap(heap_size = 10000)
outfiles = r$output_files
names(outfiles) = nii.stub(outfiles, bn = TRUE)
scheme_file = camino_fsl2scheme(
  bvecs = outfiles["bvecs"], bvals = outfiles["bvals"],
  bscale = 1)
```

The imaging scheme contains measurements at b=5, b=1000, b=2000, and b=3000 s / mm^2.

## Subsetting data
By selecting a subset of the measurements, we can reduce processing time and memory requirements. Also, the high b-value shells aren't optimal for estimating the diffusion tensor. So we'll select data from the b=5 and b=1000 shells, which is still higher angular resolution than most DTI (90 directions).

If you have sufficient RAM, you can load the whole data set and extract a subset:
```{r}
camino_ver = packageVersion("rcamino")
if (camino_ver < "0.5.2") {
  devtools::install_github("muschellij2/rcamino")
}
sub_data_list = camino_subset_max_bval(
  infile = outfiles["data"],
  schemefile = scheme_file,
  max_bval = 1500,
  verbose = TRUE) 
sub_data = sub_data_list$image
sub_scheme = sub_data_list$scheme
```


# Fit the diffusion tensor

```{r}
# wdtfit caminoProc/hcp_b5_b1000.Bfloat caminoProc/hcp_b5_b1000.scheme \
# -brainmask 100307/T1w/Diffusion/nodif_brain_mask.nii.gz -outputfile caminoProc/wdt.Bdouble
# 
mod_file = camino_modelfit(
  infile = sub_data, scheme = sub_scheme, 
  mask = outfiles["nodif_brain_mask"], 
  model = "ldt_wtd")
```


## Getting FA vlaues

```{r}
# fa -inputfile caminoProc/wdt_dt.nii.gz -outputfile caminoProc/wdt_fa.nii.gz
fa_img = camino_fa_img(
  infile = mod_file,
  header = outfiles["nodif_brain_mask"],
  retimg = FALSE)
```


### Visualizing FA images

Using `ortho2`, we can visualize these FA maps:
```{r}
ortho2(fa_img)
```

## Getting MD vlaues

```{r}
# md -inputfile caminoProc/wdt_dt.nii.gz -outputfile caminoProc/wdt_md.nii.gz
md_img = camino_md_img(
  infile = mod_file,
  header = outfiles["nodif_brain_mask"],
  retimg = FALSE)
```

# Export DTs to NIfTI

Using `camino_dt2nii`, we can export the diffusion tensors into NIfTI files.  We see the result is the filenames of the NIfTI files, and that they all exist (otherwise there'd be an errors.) 
```{r}
# dt2nii -inputfile caminoProc/wdt.Bdouble -header 100307/T1w/Diffusion/nodif_brain_mask.nii.gz \
# -outputroot caminoProc/wdt_
mod_nii = camino_dt2nii(
  infile = mod_file,
  header = outfiles["nodif_brain_mask"])
```

```{r}
# dteig -inputfile caminoProc/wdt.Bdouble -outputfile caminoProc/wdt_eig.Bdouble
eigen_image = camino_dteig(infile = mod_file)
```


We can read these DT images into `R` again using `readnii`, but we must set `drop_dim = FALSE` for diffusion tensor images because the pixel dimensions are zero and `readnii` assumes you want to drop "empty" dimensions

```{r}
dt_imgs = lapply(mod_nii, readnii, drop_dim = FALSE)
```
