---
title: "Installing ITKR on Windows"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true      
bibliography: ../refs.bib      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

All code for this document is located [here](https://github.com/muschellij2/neuroc/blob/master/itkr_windows/index.Rmd).

# Why ITKR

`ITKR` is a wrapper for ITK (https://itk.org/), which is the Insight ToolKit or now labeled Insight Segmentation and Registration Toolkit.  ITK provides a whole suite of C++ programs that allow users to perform imaging analyses and many packages rely on this, such as ANTs.  

## Why Windows



# Installing from Binary

The easiest way to install `ITKR` is through a binary file.  These have been compiled using Appveyor and a workflow using [RTools](https://cran.r-project.org/bin/windows/Rtools/).  The binary files are located at https://github.com/muschellij2/ITKR/releases and will have the extension `.zip`.  Download this file to your hard drive, and then run 

```{r, eval = FALSE}
install.packages(pkgs = "/path/to/file.zip", repos = NULL)
```
where `/path/to/file.zip` is the location of the file on your computer.  You should now have a working copy of `ITKR`. 

If for any reason, this has some sort of problem or you have a custom setup of your R installation, please go to [Compiling `ITKR` from source](#compiling-itkr-from-source).

# Compiling `ITKR` from source

Compiling `ITKR` from source is rather straightforward on OS X/Linux with standard compilers but some additional steps are required for Windows.  

You will need:
1. `git` [Download Git here](https://git-scm.com/downloads)
2. `cmake` [Download CMake here](https://cmake.org/download/)
3. `RTools` [Download RTools here](https://cran.r-project.org/bin/windows/Rtools/) (Windows Only).
4. The `devtools` or `remotes` packages (one or the other will do)
```{r, eval = FALSE}
install.packages("devtools")
install.packages("remotes")
```

## Why is it a problem on Windows?

Due to some restrictions on file path names in `CMake`, file paths for code cannot be over 50 characters long on Windows.  This issue arises when you install a package via `install_github` and the package is compiled in your temporary directory.   The file is something like `C:\tmpdir\asdfadkljwer\devtools-muschellij2-ITKR-COMMIT`, which is not a problem for other packages not using `CMake`. 

## Putting things in your PATH




**NB**: I am using my fork of ITKR from https://github.com/muschellij2/ITKR, although I am not the author of the package and the original authors package is located at https://github.com/stnava/ITKR.  This is so that I am ensured control of that repo for this downloading purpose. 


We will present 2 methods of trying to fix this: 1) building locally and 2) changing your temporary directory.
 
## Building locally 

Here we will use the Git Bash client if you have it available or any terminal 
```bash
cd C:\
git clone https://github.com/muschellij2/ITKR
R CMD INSTALL ITKR
```


## Changing your temporary directory

```bash
TMPDIR=C:\
TMP=C:\
echo $TMPDIR
echo $TMP
R
```

```{r, eval = FALSE}
remotes::install_github("muschellij2/ITKR")
```




# Session Info

```{r}
devtools::session_info()
```
