---
title: "Installing devtools"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
---

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/installing_devtools/index.R).

# First Pass

Overall, RStudio provides a fantastic tutorial and discussion on [installing devtools](https://www.rstudio.com/products/rpackages/devtools/).  Please consult this before the rest of the document.  If you have errors, please see below.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As Neuroconductor is GitHub-based, we will need a way for R to install packages directly from GitHub.  The `devtools` package provides this functionality.  In this tutorial, we will go through the steps of installing `devtools`, and some common problems.  You must have `devtools` installed to install from GitHub in subsequent tutorials on installing Neuroconductor packages.

There are other packages that will do this and are more lightweight (see `remotes` and `ghit`), but we will focus on `devtools`. 


# Mac OSX

You need to install [Command Line Tools](https://developer.apple.com/library/content/technotes/tn2339/_index.html), aka the command line tools for Xcode, if you have not already.  [http://osxdaily.com/2014/02/12/install-command-line-tools-mac-os-x/](http://osxdaily.com/2014/02/12/install-command-line-tools-mac-os-x/) is a great tutorial how.

# Installing devtools

If you already have `devtools` installed great! (Why are you in this section?)  You can always reinstall the most up to date version from the steps below.

```{r, eval = FALSE}
packages = installed.packages()
packages = packages[, "Package"]
if (!"devtools" %in% packages) {
  install.packages("devtools")
}
```

# The `remotes` and `ghit` packages
If you want a lighter-weight package that has the `install_github` functionality that `devtools` provides, but not all the "development" parts of `devtools`, the `remotes` package exists just for that:

```{r, eval = FALSE}
packages = installed.packages()
packages = packages[, "Package"]
if (!"remotes" %in% packages) {
  install.packages("remotes")
}
```

The `ghit` package is the lightest-weight package I have seen which has a `install_github` function, but may have some limited functionality compared to `remotes` in the functionality of installing package with dependencies in other systems, such as BitBucket.

In any subsequent tutorial, when you see `devtools::install_github`, just insert `remotes::install_github` and it should work just the same.


# Updating a package

In the `install_github` function, there are additional options to pass to the `install` function from `devtools`.  One of those arguments is `upgrade_dependencies`, which default is set to `TRUE`.  So if you want to install a package from GitHub, but not update any of the dependencies, then you can use `install_github(..., upgrade_dependencies = FALSE)`.  

# Troubleshooting errors 

## git2r dependency in devtools

If you cannot install `devtools`, many times it is due to `git2r`.  You should look at the installation logs and if you see something like:

```
   The OpenSSL library that is required to
   build git2r was not found.

   Please install:
libssl-dev    (package on e.g. Debian and Ubuntu)
openssl-devel (package on e.g. Fedora, CentOS and RHEL)
openssl       (Homebrew package on OS X)
```

Then run `sudo apt-get libssl-dev` or `sudo yum install openssl-devel` on your respective Linux machine.  Try to re-install `devtools`.

### Mac OSX

For Mac, you have to [install Homebrew](http://www.howtogeek.com/211541/homebrew-for-os-x-easily-installs-desktop-apps-and-terminal-utilities/) which the tutorial is located in the link.  After Homebrew is installed you should be able to type in the Terminal:
```
brew update
brew install openssl
```
Then try to re-install `devtools`.

# Session Info

```{r}
devtools::session_info()
```

