---
title: "Segmentation of Hemorrhagic Stroke in CT data"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true      
---


All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/ct_ich_seg/index.R).


```{r setup, include=FALSE}
library(ichseg)
library(dplyr)
library(readr)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, comment = "")
```


# Goal
In this tutorial, we will discuss segmentation of X-ray computed tomography (CT) scans.  The data is discussed in Validated automatic brain extraction of head CT images" (http://doi.org/10.1016/j.neuroimage.2015.03.074).  The data is located at https://archive.data.jhu.edu/dataset.xhtml?persistentId=doi:10.7281/T1/CZDPSX and was from the MISTIE (https://doi.org/10.1016/S1474-4422(16)30234-4Get) and CLEAR (https://doi.org/10.1111/ijs.12097) studies.  The MISTIE study focused on patients with intraparenchymal/intracerebral hemorrhage (ICH) and CLEAR focused on intraventricular hemorrhage (IVH), but also has patients with ICH.



## Setting up the Dataverse

The JHU archive is a [Dataverse](https://guides.dataverse.org/en/latest/) archive.  We can use the [`dataverse`](https://github.com/iqss/dataverse-client-r) package.  We will set the `DATAVERSE_SERVER` variable as this is the default variable that is used in the `dataverse` package.  I have set the environment variable `JHU_DATAVERSE_API_TOKEN` with the API token for this repository.  


```{r}
library(dataverse)
Sys.setenv("DATAVERSE_SERVER" = "archive.data.jhu.edu")
token = Sys.getenv("JHU_DATAVERSE_API_TOKEN")
```

With these set up, we can use the `dataverse` functions, by passing in `key = token` for all functions.  Alternatively, we can set:
```{r}
Sys.setenv("DATAVERSE_KEY" = Sys.getenv("JHU_DATAVERSE_API_TOKEN"))
```
and not have to set anything again.

## Finding the ID of the Dataset
Although we know the DOI is `10.7281/T1/CZDPSX` as we can see this in the URL itself https://archive.data.jhu.edu/dataset.xhtml?persistentId=doi:10.7281/T1/CZDPSX, we will use the `dataverse` functionality:
```{r}
x = dataverse_search("muschelli AND head ct")
doi = x$global_id
doi
```

## Listing the Data Files

We will get the tiles from the data set so that we can download individual files and show how to segment a specific scan.

```{r}
files = dataverse::get_dataset(doi)
files
```

We can download the demographics data from the repository so we can see some information about these patients.  We will create a wrapper function as the `get_file` function always returns a raw vector:

```{r dl_file}
library(readr)
dl_file = function(file, ...) {
  outfile = file.path(tempdir(), basename(file))
  out = get_file(file, ...)
  writeBin(out, outfile)
  return(outfile)
}
fname = grep("Demog", files$files$label, value = TRUE)
demo_file = dl_file(fname, dataset = doi)
demo = readr::read_csv(demo_file)
head(demo)
```

Here we will grab one patient, download the tarball, and then `untar` the files:

```{r download_tarball}
library(dplyr)
set.seed(20210217)
run_id = demo %>% 
  filter(dx == "ICH") %>% 
  sample_n(1) %>% 
  pull(id)
fname = paste0(run_id, ".tar.xz")
tarball = dl_file(fname, dataset = doi)
xz_files = untar(tarball, list = TRUE)
```

Here we create a temporary directory and extract the tarball to that directory.  We create a vector of the file names and extract specifically the image and the mask:
```{r extract_data}
tdir = tempfile()
dir.create(tdir)
untar(tarball, exdir = tdir)
nii_files = list.files(path = tdir, recursive = TRUE, full.names = TRUE)
nii_file = nii_files[!grepl("Mask", nii_files) & grepl(".nii.gz", nii_files)]
mask_file = nii_files[grepl("_Mask.nii.gz", nii_files)]
```

## Reading in the Data

Here we read the data into `R` into a `nifti` object:
```{r readnii, cache = FALSE}
library(neurobase)
img = readnii(nii_file)
mask = readnii(mask_file)
ortho2(img)
range(img)
```

Here we plot the image and the Winsorized version to see the brain tissue:

```{r}
ortho2(img, window = c(0, 100))
masked = window_img(mask_img(img, mask))
ortho2(masked)
```

## Segment Image

We can segment the image using `ichseg::ich_segment` to segment the image using PItcHPERFeCT (https://doi.org/10.1016/j.nicl.2017.02.007).  We will use the `ichseg::predict_deepbleed`, which implements the DeepBleed model from https://github.com/msharrock/deepbleed (https://doi.org/10.1007/s12021-020-09493-5).  We will pass in the image and the mask.  The model weights will be downloaded and the model will be run.  The `outdir` argument can be used if you would like to download the model weights to a temporary directory or if you cannot write to your R library folder. 

```{r seg}
library(ichseg)
segmentation = ichseg::predict_deepbleed(nii_file, mask_file)
print(names(segmentation))
print(segmentation)
```

We see the segmentation returns a number of things, and we want to overlay the image with the segmentation in native space (template space registered and predicted images are also included):
```{r segplot}
ortho2(masked, segmentation$native_prediction)
```

We see that some areas are dark in this image, which is because the prediction is not binary.  Areas with any values > 0 are imposed on the data.  Here we will threshold the data at a number of thresholds to show the segmentation overlaid on the image

```{r seg_thresh}
ortho2(masked, segmentation$native_prediction > 0.5, col.y = scales::alpha("red", 0.5))
ortho2(masked, segmentation$native_prediction > 0.9, col.y = scales::alpha("red", 0.5))
```

## Wrapper Function

We can create a simple function to perform the extraction and segmentation of any of the patient data in our complete data set:

```{r func}
predict_ich_data = function(id) {
  fname = paste0(id, ".tar.xz")
  tarball = dl_file(fname, dataset = doi)
  xz_files = untar(tarball, list = TRUE)
  tdir = tempfile()
  dir.create(tdir)
  untar(tarball, exdir = tdir)
  nii_files = list.files(path = tdir, recursive = TRUE, full.names = TRUE)
  nii_file = nii_files[!grepl("Mask", nii_files) & grepl(".nii.gz", nii_files)]
  mask_file = nii_files[grepl("_Mask.nii.gz", nii_files)]
  ichseg::predict_deepbleed(nii_file, mask_file)
}
```


