---
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
project_name = readLines("project_name")[1]
master_name = switch(
  project_name,
  neuroconductor = "Neuroconductor",
  oslerinhealth = "OSLERinHealth"
)
fname = "install"
if (project_name != "neuroconductor") {
  fname = paste0(fname, "_", project_name)
}
website = switch(
  project_name,
  neuroconductor = "https://neuroconductor.org",
  oslerinhealth = "http://oslerinhealth.org"
)
lite_file = switch(
  project_name,
  neuroconductor = "https://neuroconductor.org/neurocLite.R",
  oslerinhealth = "http://oslerinhealth.org/oslerLite.R"
)
installer = switch(
  project_name,
  neuroconductor = "neuro_install",
  oslerinhealth = "osler_install")
lite_installer = switch(
  project_name,
  neuroconductor = "neurocLite",
  oslerinhealth = "oslerLite")
install_pack = switch(
  project_name,
  neuroconductor = "neurocInstall",
  oslerinhealth = "oslerInstall")
packs = switch(
  project_name,
  neuroconductor = c("fslr", "neurohcp"),
  oslerinhealth = c("baker", "spotgear")
)
dpacks = paste0('c("', packs[1], '", "', packs[2], '")')
```

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/neuroc/master/`r fname`/index.R).

# Installing `r master_name` Packages 

[Install and start the latest release version of R](#installing-and-starting-r).  Although the installer will try to download and install `devtools`, there may be some system requirements for `devtools` that you may need before going forward.  Please visit [installing devtools](../installing_devtools/index.html) before going forward if you do not have `devtools` currently installed. 

Then, you can install a package using the following command:

```r
## try http:// if https:// URLs are supported
source("`r lite_file`")
`r installer`("PACKAGE")
```
where `PACKAGE` is the name of the package you'd like to install, such as ``r packs[1]``.  For example, if we want to install ``r packs[2]`` and ``r packs[1]`` we can run:
```r
source("`r lite_file`")
`r installer`(`r dpacks`)
```
### ``r lite_installer``: an alias for ``r installer``

As Bioconductor uses the `biocLite` function to install packages, we have created a duplicate of ``r installer``, called ``r lite_installer``, for ease of use for those users accustomed to Bioconductor.  The same command could have been executed as follows:
```r
source("`r lite_file`")
`r lite_installer`(`r dpacks`)
```

### Installing the ``r install_pack`` package

The ``r install_pack`` package contains the ``r lite_installer``/``r installer`` functions, as well as others relevant for `r master_name`.  You can install the package as follows:

```r
source("`r lite_file`")
`r installer`("`r install_pack`")
```

After installation, you can use `` `r install_pack`::`r lite_installer`() `` to install packages without source-ing the URL above.

## Installing `r master_name` Packages without upgrading dependencies

The ``r lite_installer``/``r installer`` functions depend on `devtools::install_github`, which will upgrade dependencies by default, which is recommended.  If you would like to install a package, but not upgrade the dependencies (missing dependencies will still be installed), you can set the `upgrade_dependencies` argument to `FALSE`:

```r
`r lite_installer`(`r dpacks`, upgrade_dependencies = FALSE)
```

# Installing and starting R 

1.  Download the most recent version of R from [https://cran.r-project.org/](https://cran.r-project.org/). There are detailed instructions on the R website as well as the specific R installation for the platform you are using, typically Linux, OSX, and Windows.

2.  Start R; we recommend using R through [RStudio](https://www.rstudio.com/).  You can start R using RStudio (Windows, OSX, Linux), typing "R" at in a terminal (Linux or OSX), or using the R application either by double-clicking on the R application (Windows and OSX).

3.  For learning R, there are many resources such as [Try-R at codeschool](http://tryr.codeschool.com/) and [DataCamp](https://www.datacamp.com/getting-started?step=2&track=r).


# Packages not available on `r master_name`

If a package is not in the `r master_name` [list of packages ](`r website`/list-packages/all), then it is not located on the [`r master_name` Github](https://github.com/`r project_name`?tab=repositories).  Therefore, when installing, you'll get the following error:

```r
Error in `r installer`(...) : 
  Package(s) PACKAGE_TRIED_TO_INSTALL are not in `r project_name`
```

Once a package is located on the list of packages, then it will be available to install. 


# Session Info

```{r, cache = FALSE}
devtools::session_info()
```

# References
